---
title: "Analytical Visualization on the History of Himalyan Expeditions"
subtitle: "INFO 526 Final Project Proposal"
author: 
  - name: "Cat-Amit"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This study analyzes Himalayan expeditions conducted from 2020 to 2024. It aims to understand the preferred routes of a particular nation and to determine how death rates for different agencies vary by time of year."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

## install.packages("tidytuesdayR")
library(tidyverse)
library(janitor)

```

## Dataset

```{r}
#| label: load-dataset
#| message: false

tuesdata <- tidytuesdayR::tt_load('2025-01-21')

#Although our two questions only use exped_tidy, we have added code to make a separate peaks_tidy in case we find out that we need variables from that dataframe later on.

exped_tidy <- tuesdata$exped_tidy
#peaks_tidy <- tuesdata$peaks_tidy

exped_tidy <- clean_names(exped_tidy)
#peaks_tidy <- clean_names(peaks_tidy)

head(exped_tidy, n=10)
#head(peaks_tidy, n=10)

```

We chose this dataset because not only did our team find the prospect of analyzing Himalayan mountaineering expeditions to be incredibly interesting, but also because the dataset contained a good balance between qualitative and quantitative data, with a variety of different data types (character, integer, date, logical, etc.). The large amount of time data inspired us to think about potential time series visualizations and the variables present allowed for diverse questions from analysis of lethality to the impact of hiker background.

## Questions

Q1: Are certain routes favored by expeditions from particular nations, and do they have disparate success rates?

Q2: Do certain Agencies have a higher number of member/personal deaths than others with respect to season/date?

## Analysis plan

Data cleanup: after using clean_names(), we anticipate that data cleanup will only require ignoring NA values in the dataset.

#### Q1 Plan:

We’ll first count, for each country, how many times each route was attempted and how many of those attempts succeeded. Then we’ll make a bar chart where:

-   **X-axis** is the route name (categorical)

-   **Y-axis** is the number of attempts (numeric)

-   **Bar fill** shows the success rate (e.g. darker = higher percentage of summits)

-   **Facets** split the chart by nation so each country has its own panel

This layout lets us instantly see which routes each country prefers (tallest bars) and how well they do on them (color intensity). We can then tweak—say, switch to a heatmap of nation vs. route with attempts as tile size and success rate as color—if we want another perspective.

The variables involved in this graph are as follows:

Existing variables:

*nation* (Categorical) - Principle nationality.\
*route1,...,route4* (Categorical) - Climbing route.\
*success1,....,success4* (Binary/Numeric) - Success on route.

New variables to be created:

*route* (categorical) - A single “long” route identifier created by pivoting *route1–route4* into one column.\
*success* (binary/numeric) - Corresponding to each *route*, the matching *success* flag (0 = fail, 1 = success).\
*attempts* (numeric) - Count of expeditions per (*nation, route*) pair.\
success*\_rate* (numeric, 0-1) - For each (*nation, route*):

$$
\text{success\_rate} = \frac{\text{total successes}}{\text{total attempts}}
$$

#### Q2 Plan:

We anticipate that the best way to answer this question is to have a graph with percent total death by agency. These graphs will be faceted by season. To account for the difference in number of traveling done by season, we plan to adverage the percent death values. So far, a bar graph is the anticipated result though we are open to trying different types of graphs. If a bar graph is the final graph, we will use a stacked bar chart that shows what portion of the total death by agency is hiker or member deaths.

The variables involved in this graph are as follows:

Existing variables:\
\
*agency* - Trekking agency\
*mdeaths* - Member deaths (per trip)\
*totmembers* - total members (per trip)\
*hdeaths* - hired personnel deaths (per trip)\
*tothired* - total hired personnel (per trip)\
*year* - year\
*season_factor* - season

New variables to be created:\
\
pmdeaths - *mdeaths*/*totmembers* - percent member deaths\
phdeaths -*hdeaths*/*tothired* -percent personnel deaths\
ptotdeaths - (*hdeaths*+ *mdeaths*)/(*tothired* + *totmembers* ) - percent total deaths

For each season, we will annotate temperature and weather conditions which will be researched from internet sources. This may require a second graph that colors based on route. If this is the case, we will also require the variable: *route1,...,route4*

## Plan of Attack

| Task Name | Assignee | Status | Due | Priority | Summary |
|----|----|----|----|----|----|
| Prepare Proposal | Cat & Amit | Done | 2025-06-10 | High Priority | Collaboratively select a dataset and define two research questions. Both will come-up with a question. |
| Proposal Peer Review | Cat & Amit | In-Progress | 2025-06-13 | High Priority | Each member will review one proposal from the assigned two proposal. Members will then review each others inputs and finalize it with mutual consent. Push the final review to respective repository as issue. |
| Proposal Peer Review Edits | Cat | To-Do | 2025-06-15 | High Priority | Make edits based on reviews from peers and have the proposal ready for the instructor to review. |
| Proposal Instructor Review Edits | Amit | To-Do | 2025-06-18 | High Priority | Make edits based on review from instructor to finalize the proposal. |
| Data Cleanup | Cat | To-Do | 2025-06-19 | High Priority | Clean the dataset (handle missing values, outliers, formatting). Consult Amit for any major decisions on data handling. |
| Creating New Variables | Amit | To-Do | 2025-06-19 | Moderate Priority | Create derived variables for analysis based on research questions. Share variable definitions with Cat for feedback. |
| Creating Draft of the Graph | Cat | To-Do | 2025-06-21 | Moderate Priority | Develop initial visualizations for the two research questions. Share drafts with Amit for input. |
| Meeting to Discuss Graph Edits | Cat & Amit | To-Do | 2025-06-21 | Low Priority | Meet to review graph drafts, discuss improvements, and agree on final visualization approach. |
| Polishing the Graph | Amit | To-Do | 2025-06-22 | Moderate Priority | Refine visualizations based on meeting feedback. Ensure clarity and alignment with research questions. |
| Write-Up of Analysis and Code | Cat | To-Do | 2025-06-25 | Moderate Priority | Write explanations for code and graphs, detailing methodology and findings. Amit reviews for clarity and accuracy. |
| Create Interactive PowerPoint | Amit | To-Do | 2025-06-25 | Moderate Priority | Design an interactive PowerPoint presentation summarizing findings. Incorporate feedback from Cat on layout and content. |
| Film Presentation | Cat & Amit | To-Do | 2025-06-27 | Low Priority | Jointly record the video presentation. Cat handles setup; Amit ensures script alignment with slides. |
| Edit Presentation | Amit | To-Do | 2025-06-29 | Low Priority | Edit the video for clarity and polish. Share with Cat for final approval before submission. |
| Final Project Submission | Cat & Amit | To-Do | 2025-06-30 | Low Priority | Make any last minute changes and submit the final project with mutual sign-off from both the team members. |
